---
title: "ggplot2 versions of simple plots"
output: 
  html_document:
    fig_caption: no
    number_sections: yes
    toc: yes
    toc_float: false
    collapsed: no
---

```{r set-options, echo=FALSE}
options(width = 105)
knitr::opts_chunk$set(dev='png', dpi=300, cache=FALSE)
pdf.options(useDingbats = TRUE)
klippy::klippy(position = c('top', 'right'))
```

# Introduction #

The **R** package `ggplot2` by Hadley Wickham provides an alternative approach to the "base" graphics in **R** for constructing plots and maps, and is inspired by Lee Wilkinson's *The Grammar of Graphics* book (Springer, 2nd Ed. 2005).  The basic premise of the *Grammar of Graphics* book, and of the underlying design of the package, is that data graphics, like a language, are built upon some basic components that are assembled or layered on top of one another.  In the case of English (e.g Garner, B., 2016, *The Chicago Guide to Grammar, Usage and Punctuation*, Univ. Chicago Press), there are eight "parts of speech" from which sentences are assembled:

- *Nouns* (e.g. *computer*, *classroom*, *data*, ...)
- *Pronouns* (e.g. *he*, *she*, *they*, ...)
- *Adjectives* (e.g. *good*, *green*, *my*, *year-to-year*, ..., including articles, e.g. *a*, *the*)
- *Verbs* (e.g. *analyze*, *write*, *discuss*, *computing*, ...)
- *Adverbs* (e.g. "-ly" words, *very*, *loudly*, *bigly*, *near*, *far*, ...)
- *Prepositions* (e.g. *on*, *in*, *to*, ...)
- *Conjunctives* (e.g. *and*, *but*, ...)
- *Interjections* (e.g. *damn*)

(but as Garner notes, those categories "aren't fully settled..." p. 18).

In the case of graphics (e.g. Wickham, H., 2016, *ggplot2 -- Elegant Graphics for Data Analysis, 2nd. Ed.*, Springer, available online from [[http://library.uoregon.edu]](http://library.uoregon.edu)) the components are:

- *Data* (e.g. in a dataframe)
- *Coordinate Systems* (the coordinate system of a plot)
- *Geoms* and their "aesthetic mappings" (the geometric objects like points, lines, polygons that represent the data)

These functions return *Layers* (made up of geometric elements) that build up the plot.  In addition, plots are composed of 

- *Stats* (or "statistical transformations", i.e. curves)
- *Scales* (that map data into the "aesthetic space", i.e. axes and legends)
- *Facets* (e.g subsets of data in individual plots)
- *Legends* and
- *Themes* (that control things like background colors, fonts, etc.)

```{r source, echo=FALSE}
load(".RData")
```
# Univariate and bivariate plots # 

## Simple plots ##

Load the `ggplot2` package:

```{r load ggplot2}
# ggplot2 versions of plots
library(ggplot2)
```

Here is the **R** base graphics simple index plot:

```{r index}
# index plot
attach(sumcr)
plot(Length)
```

The `ggplot2` package contains a method, invoked by the `qplot()` function for creating quick (hence the name) versions of the base-graphics plots.

```{r index_gg}
qplot(seq(1:length(Length)), Length)
```

Here's the stripchart, plain and stacked:

```{r stripcharts}
# stripchart
stripchart(Length)
stripchart(Length, method="stack")
```

... and the `qplot()` versions:

```{r stripcharts_gg1}
qplot(Length, rep(1,length(Length)))
qplot(Length, rep(1,length(Length)), geom = "jitter") + ylim(0,4)
```

Note the use of the `geom = "jitter" arguement to get some separation among the points.

The full `ggplot2()` almost reproduces the base-graphics version.  Note the use of the `aes()` (aesthetic) function for describing the basic plot, which as the dotplot added using the `geom_dotplot()` function.  The `theme_bw()` term specifies the simple black-and-white theme.  As in the case of a histogram, the `binwidth` parameter controls the granularity of the plot.

```{r stripchargs_gg2}
ggplot(data=sumcr, aes(x = Length)) + geom_dotplot(binwidth = 0.25, method = "histodot")
```
## Dotcharts ##

Here's the base-graphics Cleveland-style dotchart.

```{r dotchart}
# Dot charts
dotchart(WidthWS, labels=as.character(Location), cex=0.5)
```

... and the `ggplot2` equivalent:

```{r dotchart_gg}
ggplot(sumcr, aes(x=WidthWS, y=Location)) + geom_point(size=1) 
ggplot(sumcr, aes(x=WidthWS, y=Location)) + geom_point(size=1) + theme_bw()
```

Note the subtle difference produced by the use of the `theme_bw()` term.  The plots are basically x-y plots (specified by the `aes(x=WidthWS, y=Location)` arguements), with the points themselves added by the `geom_point()` term.

Here's the dotchart, with rows ordered by the `WidthWS` values:

```{r dotchart_ordered}
index <- order(WidthWS)
dotchart(WidthWS[index], labels=as.character(Location[index]), cex=0.5)
```

... and the `ggplot2` equivalent--note the use of the `theme()` term to modify the label size:

```{r dotchart_ordered_gg}
ggplot(sumcr, aes(x=WidthWS, y=reorder(Location, WidthWS))) + geom_point(size=1) + 
  theme(axis.text.y = element_text(size=4, face="italic"))
```

## Boxplots ##

Here is the standard boxplot:

```{r boxplot}
# Boxplot
boxplot(WidthWS ~ HU, range=0)
```

... and the `qplot()` version:  

```{r boxplot_q}
qplot(HU, WidthWS, data=sumcr, geom="boxplot")
```

```{r boxplot_gg}
ggplot(sumcr, aes(x=HU, y=WidthWS)) + geom_boxplot()
```

The actual data points can be added to the boxplot, by specifying an additional graphical element using the `geom_point()` term.

```{r boxplot_gg2}
ggplot(sumcr, aes(x=HU, y=WidthWS)) + geom_boxplot() +
  geom_point(colour = "blue")
```
Detach the `sumcr` dataframe.

```{r detach1}
detach(sumcr)
```

## Histograms and density plots ##

The base-graphics histogram, with `breaks = 20` looks like this:

```{r histogram}
# histograms
attach(scanvote)
hist(Yes, breaks=20)
```

Here's the `ggplot2` version, with the biwidth specified explicitly (e.g., 1.0)

```{r hist_gg1}
ggplot(scanvote, aes(x=Yes)) + geom_histogram(binwidth = 1)
```

Here's another `ggplot` version, with more control over appearance, and with the edge of the first bin explicitly specified.

```{r histogram_gg2}
ggplot(scanvote, aes(x=Yes)) + geom_histogram(binwidth=1, fill="white", color="red", boundary=25.5)
```

Here's a version with the individual data points added--not how it is built up from a basic plot by progressively adding graphical elements:

```{r histogram_gg4}
ggplot(scanvote, aes(x=Yes)) + 
  geom_histogram(binwidth=1, fill="white", color="red", boundary=25.5) + geom_point(y=0) 
```

Here's a base-graphics density plot:

```{r density}
# density plots
Yes.density <- density(Yes)
plot(Yes.density)
```
... and two versions of the `ggplot2` density plot--not the subtle difference in how the density curve is drawn:

```{r density_gg}
ggplot(scanvote, aes(x=Yes)) + geom_density()
ggplot(scanvote, aes(x=Yes)) + geom_line(stat="density")
```
The first version draws the density curve as if it was a polygon, while the second looks more as expected.

The combined histogram, density and rug plot via the base graphics:

```{r histdensrug}
Yes.density <- density(Yes)
hist(Yes, breaks=20, probability=TRUE)
lines(Yes.density)
rug(Yes)
```

And here's a `ggplot` version, with essentially three distinct layers:

```{r histdenrug_gg}
ggplot(scanvote, aes(x=Yes)) +
  geom_histogram(data=scanvote, aes(x=Yes, y=..density..), binwidth=1, fill="white", color="red", boundary=25.5) + 
  geom_line(stat = "density") +
  geom_rug(data=scanvote, aes(x=Yes)) 
```
geom_density(data=scanvote, aes(x=Yes, y=..density..)) +

```{r detach3}
detach(scanvote)
```